% sec_abstract.tex — Abstract for arXiv submission
% Source: paper-abstract.md (converted 2026-05-04)
% Language: English (paper language)

Production memory systems fail not because retrieval is hard, but because silent
architectural degradation accumulates faster than any embedding model can compensate.
GraphRAG, MemGPT, Mem0, and A-MEM encode structure and recency; none encodes incident
severity, none enforces ranking-change validation. NOX-Supermem's design is
incident-driven: each failure became a schema constraint. Three contributions: (1)
Pain-weighted salience
($\text{salience} = \text{recency} \times \text{pain} \times \text{importance}$,
$\text{pain} \in [0.1, 1.0]$) as a first-class retrieval dimension, calibrated to
PagerDuty P1--P5, not psychometric scales. Ablation
($n=31$, hybrid mode) shows non-significant aggregate effect ($\Delta = +0.0065$, 95\%~CI
$[-0.014, +0.034]$); lift observed in 1/31 queries (Q55, $\Delta = +0.349$); 29/31
unaffected because semantic scores were not tied. The binding constraint
is BM25 recall ceiling: 55 of 60 queries fail
FTS-only regardless of pain calibration; full hybrid retains Recall@10 $= 0.7667$ via
Gemini. (2) Shadow discipline: a seven-day gate before any
ranking change activates, enforced via \texttt{/api/health} as an architectural
constraint, not a convention \cite{kohavi2020trustworthy,chapelle2012interleaved}. (3)
Shared-canonical context: six agents, one corpus, no federation. On 61,257 chunks, hybrid
retrieval (FTS5 + Gemini 3072-d + RRF, $k=60$) achieves
nDCG@10 $= 0.5831 \pm 0.0046$ ($n=60$, 3-run mean, post-cure gold standard v1.1) versus 0.0000 for FTS5 vanilla and
0.1475 for BM25 Pyserini ($n=60$), a $4.0\times$ lift over the strongest lexical baseline.
Edge-type enum coverage improved $14\% \to 56\%$ ($4\times$ gain, $n=100$, 95\%~Wilson
CI $[46$--$66\%]$); self-reported enum coverage rate, not human-validated accuracy. On
LOCOMO (Maharana et al., 2024, $n=100$), FTS5 achieves nDCG@10 $= 0.281$, confirming
lexical difficulty is corpus-dependent; on BEIR TREC-COVID, multilingual-e5-base achieves
nDCG@10 $= 0.8335$ ($n=50$). Open gap: corpus scale $>$100K. Code,
harness, golden set ($n=60$), and incident log:
\url{https://github.com/totobusnello/memoria-nox} (MIT). Operational discipline is at
least as important as embedding sophistication.
